Synthesizing Quotient Lenses

### 2017-10-26 papers/QuotientLensSynth/main.tex
Added squash to example. Need to make QRE section better.

\begin{abstract} 
{\em Lenses}~\cite{Lenses} specify bidirectional 
transformations between different data formats.
\saz{The old abstract read more like an introduction, so we modified it to be
  the ``new'' introduction.}
\end{abstract}

\section{Introduction}
A {\em lens} $\ell$ is a {\em bidirectional transformation} between a source
type $S$ and a target type $T$ in which the forward transformation
$\ell.\get$ and the backward transformation $\ell.\lput$ satisfy certain
round-tripping laws such as invertibility.
 Writing a single lens
rather than two separate uni-directional transformations saves both
keystrokes and bugs, by reducing redundancy and by automatically
guaranteeing strong ``lens laws'' ensuring that the forward and backward
transformations interact well.

@@

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{Boomerang} is the only implementation
currently offering quotient lenses.  One reason for this is that there is
some annoying redundancy in the quotient lens notation: The programmer must
separately specify (1) the data formats on each side of the lens, (2) the
equivalence relations over them, and (3) a lens between the canonical
representatives of the equivalence classes, even though, typically, these
three share a great deal of structure.  Moreover, writing the lens between
the equivalence classes can be challenging due to the rather fiddly, mainly
due to ``non-ambiguity'' conditions in the lens well-formedness rules.

This paper addresses both of these issues.  First, in Section~\ref{example} we
explain the difficulties of programming with quotient lenses and show, by way of
example, how our novel language of {\em Quotient Regular Expressions}, or QREs
addresses those difficulties.  Quotient regular expressions allow programmers to
specify a broad class of equivalence relations over regular languages using a
compact and convenient notation in which the regular language and its
equivalence classes are captured by a single expression.  We then define {\em
  QRE lenses}, a class of bijective quotient lenses that are typed using QREs,
and we show how to {\em synthesize} QRE lenses from their QRE types and a set of
input-output examples.  We demonstrate the practicality of this approach by
synthesizing QRE lenses between several real-world data formats from the {\tt
  data.gov} database.

@@

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a subset of bijective quotient
lenses whose types are given by QREs.  Our main technical contribution is a
normal form for QRE lenses and a proof that normal forms are
closed under left quotienting, right quotienting, composition, and regular
operators (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

\section{Lenses and Quotient Regular Expressions by Example}

An example of a bidirectional transformation which we shall revisit
severally in this paper is a lens which converts citation records such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

@@

Lenses have the advantage of saving the programmer from having to
write two seperate transformations rather than one. Also, lenses offer strong
guarantees about how the two transformations interact since lenses are
constrained by lens laws.

However, the constraints imposed by the lens laws are often too strict. For
example, when a record may be represented in many ways, perhaps because
whitespace characters are allowed to occur arbitrarily in certain parts of the
data, deriving a lens which obeys the lens laws in such cases can be daunting
and frustrating.

As a solution to this problem, Foster et al proposed {\em quotient lenses},
which are a generalization of lenses in which the source and target data are
considered modulo an equivalence relation. Quotient lenses, which are
implemented as a refinement of the bidirectional string processing language
Boomerang, are both simpler and more flexible, since whole equivalence classes
of data values can be handled all at once and inessential details like
whitespace and ordering of fields can be ignored.

However many complicated equivalence relations are difficult to express in
Boomerang. For example, suppose that the \bibtex{} to EndNote
transformation which we introduced is required to allow for whitespace to occur
arbitrarily between the words in the label, author and title fields as well as
in between them, and ignore the order in which the author and title fields
occur in each record.
These rules allow for the \bibtex{} citations
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }
\end{verbatim}
and
\begin{verbatim}
    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "J. H. Conway",     }
\end{verbatim}
to be considered the same, with the EndNote records having similar rules
defined on them.

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is difficult. The difficulty arises from the fact
that the entire equivalence is composed of three separate equivalences which
nest and interact with each other in a way that makes them difficult to express
with the standard Boomerang combinators. Moreover, many of the equivalences
expressible in Boomerang, such as Boomerang's permutation combinator, require
that the programmer first specify the data itself in form of a regular
expression before specifying the equivalence relation on the data. This leads to
verbose code that is difficult to understand.

To address this issue, we propose the language of {\em Quotient Regular
Expressions} or QREs, which express a broad class of equivalence relations
directly on regular languages using a compact and convenient notation. For
example, with QREs we can specify \bibtex{} records as shown in
Figure~\ref{fig:example-qre}.

\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]$^+$
let wsp_to_space = project wsp $\to$ " "
let valid_name = name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let wsp_to_and = project (wsp . "and" . wsp) $\to$ " and "
let bib_names = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title = "title = {" . title . "}"
let bib_perm = perm (label, bib_author, bib_title) with (project ("," .wsp) $\to$ ",\n")
let bibtex = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

The expression \cd{project [\n\t\r]}$^+$ $\to$ \cd{" "} is a QRE that matches any 
number of whitespace characters, and then considers all strings of whitespace
characters as equivalent with a single space as the canonical representative. 
Another QRE in the example above is the

\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . wsp ) $\to$ ",\n")
\end{lstlisting}
\end{center}

@@

While QREs make it easier to write quotient lenses by specifying the data and
the equivalence relation on the data in one expression, the programmer still
has to manually derive the underlying lens which maps the equivalence classes
of the source data to the equivalence classes of the target data. This is
itself tedious and error-prone task, as the code snippet below illustrates:

@@

QRE lenses lenses are the same as bijective lenses with ``canonizers at the
edges'' as in Figure~\ref{fig:attheedges}, and are typed using QREs. In the
forward direction, a QRE lens $q$ uses the source QRE $c$ to compute a unique
representative for the data modulo the equivalence relation defined by $c$ and
then applies the $\get$ function of a bijective lens $\ell$ to this
representative. In the backward direction, $q$ operates similarly,  but using
the view QRE $c'$ and the $\lput$ function of $\ell$.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{canonizers-outside}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
      screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
      (if we put ``from [Nate's thesis]'' in the caption), but I don't think
      we want the full complexity of asymmetric lenses here.  We should draw
      the figure again.  (I have the OmniGraffle file for it, so this is
      easy if I know how it should look.)}\bcp{Done---what do you think?}
      \sam{Yes this is more accurate, except that we don't need a general choose
      function. With QREs the kernel language is always contained in the whole
      language so we just use the inclusion as the choose function. Also,
      perhaps it would be clearer if we said source structures and target
      structres instead of all structures?}}
  \label{fig:attheedges}
\end{figure}

One potential pitfall with QRE lenses is that the combinators which
we define on them may not preserve this normal form. For example, the
composition $q' \circ q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens which has the canonical normal form of a bijective lens with
canonizers a the edges. Our main technical contribution in this paper is a
proof that QRE lenses are closed under left quotienting, right quotienting,
composition, and regular combinators.\bcp{If this is our main technical
  contribution, then it deserves more than a sentence in the intro!  E.g.,
  can we give an example?} This proof will allow justify our
approach for synthesizing quotient lenses so that the programmer can simply
type

\begin{lstlisting}
let bib_to_end : (lens in bibtex <=> endnote) 
= synth bibtex <=> endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent to synthesize a quotient lens that maps between \cd{bibtex} and
\cd{endnote} QREs, and sends the equivalence class of \cd{bib_bexample} to
the equivalence class of \cd{end_example}.

>>>>

\begin{abstract} 
\saz{The old abstract read more like an introduction, so we modified it to be
  the ``new'' introduction.}
\end{abstract}

\section{Introduction}
A {\em lens} $\ell$ is a {\em bidirectional transformation} between a source
type $S$ and a target type $T$ in which the forward transformation
$\ell.\get$ and the backward transformation $\ell.\lput$ satisfy certain
round-tripping laws such as invertibility. Lenses are advantageous
because writing a single lens rather than two separate uni-directional
transformations saves both keystrokes and bugs, by reducing redundancy and by
automatically guaranteeing strong ``lens laws'' ensuring that the forward and
backward transformations interact well.

@@

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{Boomerang} is the only implementation
currently offering quotient lenses.  One reason for this is that there is
some annoying redundancy in the quotient lens notation: The programmer often
has to separately specify (1) the data formats on each side of the lens, (2) the
equivalence relations over them, and (3) a lens between the canonical
representatives of the equivalence classes, even though, typically, these three
share a great deal of structure.  Moreover, writing the lens between the
equivalence classes can be challenging due to the rather fiddly, mainly due to
``non-ambiguity'' conditions in the lens well-formedness rules.

This paper addresses both of these issues.  First, in Section~\ref{example} we
explain the difficulties of programming with quotient lenses and show, by way of
example, how our novel language of {\em Quotient Regular Expressions}, or QREs
and QRE lenses address these difficulties.  Quotient regular expressions allow
programmers to specify a broad class of equivalence relations over regular
languages using a compact and convenient notation in which the regular language
and its equivalence classes are captured by a single expression.  We then
define {\em QRE lenses}, a class of bijective quotient lenses that are typed
using QREs, and we show how to {\em synthesize} QRE lenses from their QRE types
and a set of input-output examples.  We demonstrate the practicality of this
approach by synthesizing QRE lenses between several real-world data formats
from the {\tt data.gov} database.

@@

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a subset of bijective quotient
lenses whose types are given by QREs. Our main technical contribution in this
paper is a proof that each QRE lenses $q$ defines the same transformation as a
QRE lens $q'$ which is in normal form. (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

\section{Quotient Regular Expressions and Lenses by Example}
Consider a bidirectional transformation which converts citation records such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

@@

We wish for this transformation to have the same behaviour on records which
differ only in ``inessential'' details. First, the transformation
should allow whitespace characters to occur arbitrarily between the words in the
label, author and title fields as well as in between them. Second, an
author's name can have the surname followed by a comma and then the other
names separated by arbitrary whitespace, or just all the names separated by
arbitrary whitespace (for example, the names ``Conway, J. H.'' and ``J. H.
Conway'' are equivalent). Lastly, the transformation should also ignore the
order in which the author and title fields occur in each citation record.

For example, with these rules, the following \bibtex{} citations should be
considered the same by the lens,
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }

    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "Conway, J. H.",     }
\end{verbatim}
with the EndNote records having similar rules defined on them.

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge. This is because the entire
equivalence is composed of three separate equivalences which nest and interact
with each other in a way that makes them difficult to express with the standard
Boomerang combinators. Moreover, many of the equivalences expressible in
Boomerang require that the programmer first specify the data itself in form of
a regular expression before specifying the equivalence relation on the data.
This leads to verbose, repetitive code.

To address this issue, we propose {\em Quotient Regular Expressions} or QREs,
which express a broad class of equivalence relations directly on regular
languages using a compact and convenient notation. For example, with QREs we
can specify \bibtex{} records as shown in Figure~\ref{fig:example-qre}.

\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]
let wsp_to_space = project wsp+ $\to$ " "
let wsp_to_space_epsilon = project wsp* $\to$ ""
let space_name  = name | (name . wsp_to_space . (name . wsp_to_space)* . name)
let wsp_to_comma = wsp_to_space_epsilon . "," . wsp_to_space
let comma_name = name | (name . wsp_to_comma . (name . wsp_to_space)* . name)
let space_to_comma = synth comma_name <=> space_name using {("Conway, J. H.", "J. H. Conway")}
let valid_name = squash comma_name -> space_name using (get space_to_comma)
let wsp_to_and = project (wsp+ . "and" . wsp+) $\to$ " and "
let bib_names = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title = "title = {" . title . "}"
let bib_perm = perm (label, bib_author, bib_title) with (project ("," .wsp+)
$\to$ ",\n") let bibtex = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

The expression \cd{project [ \\n\\t\\r]}$^+$ $\to$ \cd{" "} is a QRE that
matches any number of whitespace characters, and then considers all strings of
whitespace characters as equivalent with a single space as the canonical
representative. Another QRE in the example above is the expression

\begin{center}
\begin{lstlisting}
squash comma_name -> space_name using (get space_to_comma)
\end{lstlisting}
\end{center}

\noindent which considers the \cd{comma_name} representation of the author's
name to be equivalent to the \cd{space_name} representation, with \cd{get
space_to_comma} giving the function that transforms names in the \cd{comma_name}
representation to names in the \cd{space_name} representation.

The third example of a QRE in the example is the expression
\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . wsp ) $\to$ ",\n")
\end{lstlisting}
\end{center}

@@

Now QREs make it easier to write quotient lenses by specifying the data and
the equivalence relation on the data in one expression, but the programmer still
has to manually derive the underlying lens which maps the equivalence classes
of the source data to the equivalence classes of the target data. This is
itself a tedious and error-prone task, as the code snippet below illustrates:

@@

QRE lenses lenses are the same as bijective lenses with ``canonizers at the
edges'' as in Figure~\ref{fig:attheedges}, and are typed using QREs - every QRE
lens $q$ has a type $q: c \Leftrightarrow c'$ where $c$ and $c'$ are QREs.
In the forward direction, a QRE lens $q: c \Leftrightarrow c'$ uses the source
QRE $c$ to compute a unique representative for the data modulo the equivalence
relation defined by $c$ and then applies the $\get$ function of a bijective
lens $\ell$ to this representative. In the backward direction, $q$ operates
similarly,  but using the view QRE $c'$ and the $\lput$ function of $\ell$.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{canonizers-outside}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
      screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
      (if we put ``from [Nate's thesis]'' in the caption), but I don't think
      we want the full complexity of asymmetric lenses here.  We should draw
      the figure again.  (I have the OmniGraffle file for it, so this is
      easy if I know how it should look.)}\bcp{Done---what do you think?}
      \sam{Yes this is more accurate, except that we don't need a general choose
      function. With QREs the kernel language is always contained in the whole
      language so we just use the inclusion as the choose function.}}
  \label{fig:attheedges}
\end{figure}

One potential pitfall with QRE lenses is that the combinators which
we define on them may not preserve this normal form. For example, the
composition $q' \circ q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens which has the canonical normal form of a bijective lens with
canonizers a the edges. Our main technical contribution in this paper is a
proof that each QRE lenses $q$ defines the same transformation as a QRE lens
$q'$ which is in normal form.

For example, suppose \cd{q_1 : bibtex <=> endnote} is a QRE lens that maps
between \bibtex{} and EndNote formats, and that \cd{q_2 : endnote <=> xml}
is a QRE lens that maps between EndNote and XML formats. The composition 
\cd{(q_1 ; q2) : bibtex <=> xml} of \cd{q_1} and \cd{q_2} is not in normal form
due to the presence of canonizers on each side of the intermediary EndNote
format:

\sam{TODO: Put diagram of composition here}

\noindent However, while \cd{q_1 ; q_2} is not in normal form, there is another
QRE lens \cd{q : bibtex <=> xml} of the same type as \cd{q_1 ; q_2} such that
\cd{q_1 ; q_2} and \cd{q} define the same transformation. This proof will allow
justify our approach for synthesizing quotient lenses so that the programmer
can simply type

\begin{lstlisting}
let bib_to_end : (lens in bibtex <=> endnote) 
= synth bibtex <=> endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent to synthesize a quotient lens that maps between the \cd{bibtex} and
\cd{endnote} QREs, and sends the equivalence class of \cd{bib_bexample} to
the equivalence class of \cd{end_example}.
